---
title: "LongitudinalDataAnalysis"
author: "Group2: Wanchang Zhang; Hugo Blain; Oscar Cabanelas"
date: "2023-03-05"
output: pdf_document
---
# 0. Introduction 
The dataset contains information on patients who received renal graft(kidney transplant)
The patients have been followed for at most 10 years.

Background: People with end-stage kidney disease who receive a kidney transplant generally live longer than people with ESRD who are on dialysis. 

However, kidney transplant recipients must remain on immunosuppressants  (medications to suppress the immune system) for the rest of their life to prevent their body from rejecting the new kidney. The long-term immunosuppression puts them at risk for infections and cancer.

The Haermatocrit level (HC level) usually differs with gender, Also the health condition of a person.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# 1. Task for week one
## 1.1 Import data
```{r}
#install.packages("readxl")
library("readxl")
trenal <- read_excel("Trenal.XLS")
summary(trenal)
```
remove a noninformative column const
```{r}
trenal= trenal[,-18]
summary(trenal)
```
```{r}
dim(trenal)
``` 
## 1.2 Table structure analysis and variable understanding
The table contains observation of HC level on 1160 patients who have gone through kidney transplant. Each patient will have maximum 12 meassurements in the 12 time point $(0,0.5,1,2,\cdots,10)$ years.


if we just look at the first 12 columns, they are all Haematocrit level at the corresponding time. Thus our response variable is Haematocrit level. If we just look at first 17 columns from HC0 to reject, then the subtable looks like a wide table;
If we start from column id to column time, the part of table is a long table.
From now on we focus on the long table:
```{r}
trenal.long = trenal[,13:20]
summary(trenal.long)
```


```{r}
dim(trenal.long)
```
Besides the time $0,0.5,1,2,3,4,5,\cdots,10$ is one-to-one correspondent to j $1,2,3,\cdots,12$. But we can still leave it in the dataframe.
Our response variable is the HC level(The percentage of red cells in the blood, normal levels of hermatocrit for men range from $41\%$ to $50\%$, normal level for women is $36\%$ to $48\%$)
the explanatory variables are age, 
we can change the structure of the table as we are used to: Identity, time, respons, explanatory variables (time dependent), explanatory variables (time independent).
The response variables are some continuous integer values? 
The explanatory variables have binary type: male, cardio, reject,
and integer type: age
```{r}
#install.packages("magrittr") # package installations are only needed the first time you use it
#install.packages("dplyr")    # alternative installation of the %>%
library(magrittr) # needs to be run every time you start R and want to use %>%
library(dplyr)   
data <- trenal.long %>% 
  relocate(id) %>%
  relocate(j,.after=id)%>%
  relocate(time,.after = j)%>%
  relocate(respons,.after=time)
trenal.long$id = as.factor(trenal.long$id)
trenal.long$j = as.factor(trenal.long$j)
trenal.long$male = as.factor(trenal.long$male)
trenal.long$cardio = as.factor(trenal.long$cardio)
trenal.long$reject = as.factor(trenal.long$reject)
summary(trenal.long)
length(unique(trenal.long$id))
```

```{r}
# Plot the raw data
plot(trenal.long$time,trenal.long$respons)
```
```{r}
library(ggplot2)
library(nlme)
library(lme4)

#Plot data
ggplot(data, aes(x=time, y=respons)) + geom_point()

#Plot data with lm line
ggplot(data, aes(x=time, y=respons)) + geom_point() + geom_smooth(method="lm")

```


## 1.3 List of Hypotheses to be tested by the data
```{r}
#Select a sample of data to plot
set.seed(1)
selected <- sample(1:length(unique(trenal.long$id)),30,replace=T) # random samples and permutations
#selected.vector = as.vector(selected)
data.selected = data[(data$id %in% c(selected)), ] 
```

#### Spaghetti Plot the response over time with the different persons
```{r}

#data.selected = data[data$id == selected.vector,]# why the dim(data.selected) = 12 x 8
# Plot the respons over time for different id
# If some responses are not available, NA
ggplot(data.selected, aes(x=time, y=respons, group=id,color=id)) + geom_point()  +geom_line()
#ggplot(data.selected, aes(x=time, y=na.pass(respons), group=id,color=id)) + geom_point()  +geom_line()
```

##### Bar plot response over time
```{r}
# Box plot
p <- ggplot(data, aes(x=time, y=respons,group =time, color = time)) +  
  geom_boxplot()
p
```
#### Hypothese one
HC level will change with time differently if the REJECT is different
```{r}
#Plot individual data
ggplot(data, aes(x=time, y=respons, group=reject,color=reject)) + geom_point() 
```
##### Bar plot 

```{r}
# Box plot
p <- ggplot(data, aes(x=time, y=respons,group = reject, color = reject)) +  
  geom_boxplot()
p
```


#### Hypothese two
HC level will change with time differently if the gender is different, male (1) has generally higher HC level than female (0)
```{r}
data.male =  data[(data$male == "1" ), ] 
data.female =  data[(data$male == "0" ), ] 
data.male.selected =  data[(data$male == "1" & data$id %in% c(selected)), ] 
data.female.selected =  data[(data$male == "0" & data$id %in% c(selected)), ]
```

#####  Spaghetti plots stratified by variables gender male
```{r}
ggplot(data.male.selected, aes(x=time, y =respons, group=id,color=id)) + geom_point()  +geom_line()+ggtitle("Spaghetti plot for male")+ geom_smooth(method="lm")
ggplot(data.female.selected, aes(x=time, y= respons, group=id,color=id)) + geom_point()  +geom_line()+ggtitle("Spaghetti plot for female")+ geom_smooth(method="lm")
```
##### Box plot
```{r}
p <- ggplot(data, aes(x=male, y=respons,group = male, color = male)) +  
  geom_boxplot()
p
```

##### Box plot for male and female
COULD I PLOT THEM IN THE SAME FIGURE?
```{r}

p.box.male <- ggplot(data.male, aes(x=time, y=respons,group = time, color = time)) +  
  geom_boxplot()
p.box.male

p.box.female <- ggplot(data.female, aes(x=time, y=respons,group = time, color = time)) +  
  geom_boxplot()
p.box.female
```



#### Hypothese three
HC level will change with time differently if the age when performing the kidney transplant is younger
```{r}
#Plot individual data
ggplot(data.selected, aes(x=time, y=respons, group=age,color=age)) + geom_point() +geom_line()
```

##### Box plot
```{r}
p <- ggplot(data, aes(x=age, y=respons,group = age, color = age)) +  
  geom_boxplot()
p
```
#### Hypothese four
HC level will change with time differently if the patient has experienced cardio-vascular problem during the years preceding the tranplantation
##### Spathetti plot ?
```{r}
#Plot individual data
ggplot(data.selected, aes(x=time, y=respons, group=cardio,color=cardio)) + geom_point()  +geom_line()
```
##### Bar plot
```{r}
p <- ggplot(data, aes(x=cardio, y=respons,group = cardio, color = cardio)) +  
  geom_boxplot()
p
```

## Correlation analysis for different HC levels along time
For this purpose we need the wide table

```{r}
trenal.wide = trenal[,1:17]
summary(trenal.wide)
```
```{r}
cor(trenal.wide$HC0,trenal.wide$HC06)
```
```{r}
# scatter plot matrix

pairs(~HC0+HC06+HC1+HC2+HC3+HC4+HC5+HC6+HC7+HC8+HC9+HC10,data=trenal.wide,
   main="Simple Scatterplot Matrix")
```



```{r}
#Lm
lm<-lm(respons~time,data=data)
summary(lm)

```

```{r}
#Plot individual data
ggplot(data, aes(x=time, y=respons, group=male,color=male)) + geom_point() 
#Spaghetti Plot
ggplot(data, aes(x=time, y=respons, group=male,color=male)) + geom_point() +geom_line()
#Spaghetti with fitted lines
ggplot(data, aes(x=time, y=respons, group=male,color=male)) + geom_point()+ geom_smooth(method="lm",se=F) +geom_line()


```
## Linear mixed effect model
Fixed effect could be time, gender, age, reject, cardio 
Random effect could be 

```{r}
#lme
#data = trenal.long
#lme <- lme(repsons ~ time + age ,data=data)
#lme<-lme(respons~time+age+male+reject+cardio,data=data)
#summary(lme)

#newdata<-data.frame(ID=c(1,2,3,4,5),week=c(3,3,3,3,3))
#newdata$prediction<-predict(lm,newdata=newdata)
#newdata
#predict(lme,newdata=newdata,level=0:1)


```


















