---
title: "Longitudinal Data Analysis "
subtitle: "Case study of Trenal.XLS using Linear Mixed Effect Model"
author: "Group2: Wanchang Zhang; Hugo Blain; Oscar Cabanelas"
date: "`r Sys.Date()`"
output: 
  pdf_document:
  fig_caption: true
  number_sections: true
---
\newpage 
\tableofcontents 
\listoffigures
\listoftables
\newpage

# 0. Introduction 
The dataset contains information on patients who received renal graft(kidney transplant)
The patients have been followed for at most 10 years.

Background: People with end-stage kidney disease who receive a kidney transplant generally live longer than people with ESRD who are on dialysis. 

However, kidney transplant recipients must remain on immunosuppressants  (medications to suppress the immune system) for the rest of their life to prevent their body from rejecting the new kidney. The long-term immunosuppression puts them at risk for infections and cancer.

The Haermatocrit level (HC level) usually differs with gender, Also the health condition of a person.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# 1. Task for week one
## 1.1 Import data
```{r}
#install.packages("readxl")
library(readxl)
trenal <- read_excel("Trenal.XLS")
summary(trenal)
```
remove a noninformative column const
```{r}
trenal= trenal[,-18]
summary(trenal)
```
```{r}
dim(trenal)
``` 
## 1.2 Table structure analysis and variable understanding
The table contains observation of HC level on 1160 patients who have gone through kidney transplant. Each patient will have maximum 12 meassurements in the 12 time point $(0,0.5,1,2,\cdots,10)$ years.


if we just look at the first 12 columns, they are all Haematocrit level at the corresponding time. Thus our response variable is Haematocrit level. If we just look at first 17 columns from HC0 to reject, then the subtable looks like a wide table;
If we start from column id to column time, the part of table is a long table.
From now on we focus on the long table:
```{r}
trenal.long = trenal[,13:20]
summary(trenal.long)
```


```{r}
dim(trenal.long)
```
Besides the time $0,0.5,1,2,3,4,5,\cdots,10$ is one-to-one correspondent to j $1,2,3,\cdots,12$. But we can still leave it in the dataframe.
Our response variable is the HC level(The percentage of red cells in the blood, normal levels of hermatocrit for men range from $41\%$ to $50\%$, normal level for women is $36\%$ to $48\%$)
the explanatory variables are age, 
we can change the structure of the table as we are used to: Identity, time, respons, explanatory variables (time dependent), explanatory variables (time independent).
The response variables are some continuous integer values? 
The explanatory variables have binary type: male, cardio, reject,
and integer type: age
```{r}
#install.packages("magrittr") # package installations are only needed the first time you use it
#install.packages("dplyr")    # alternative installation of the %>%
library(magrittr) # needs to be run every time you start R and want to use %>%
library(dplyr)   
trenal.long$id = as.factor(trenal.long$id)
trenal.long$j = as.factor(trenal.long$j)
trenal.long$male = as.factor(trenal.long$male)
trenal.long$cardio = as.factor(trenal.long$cardio)
trenal.long$reject = as.factor(trenal.long$reject)

data <- trenal.long %>% 
  relocate(id) %>%
  relocate(j,.after=id)%>%
  relocate(time,.after = j)%>%
  relocate(respons,.after=time)

summary(data)
sum(!is.na(data$respons))

length(unique(data$id))
```

```{r}
# Plot the raw data
#install.packages("tigerstats") 
require(tigerstats)
xyplot(respons ~ time, groups = id, data=data, type="l",xlab="Time (year)",ylab="HC level " )
```


```{r}
library(ggplot2)
library(nlme)
library(lme4)

#Plot data
ggplot(data, aes(x=time, y=respons)) + geom_point()

#Plot data with lm line
ggplot(data, aes(x=time, y=respons)) + geom_point() + geom_smooth(method="lm")

```


## 1.3 List of Hypotheses to be tested by the data
```{r}
#Select a sample of data to plot
set.seed(1)
selected <- sample(1:length(unique(trenal.long$id)),30,replace=T) # random samples and permutations
#selected.vector = as.vector(selected)
data.selected = data[(data$id %in% c(selected)), ] 
```

```{r}
# Individual plots
plot(data.selected) # WHAT I WILL GET FROM THE PLOT(DATA), HOW to plot a scatter plot of HC level changes with time for each id instead of a scatter plot
```
```{r}
# spaghettic plot
ggplot(data.selected,aes(x=time,y=respons,group=id,color=id))+geom_point()+ geom_line()

```

```{r}
# Plot individual data by sex
ggplot(data.selected,aes(x=time,y= respons,group=id, color=male))+geom_point()+geom_line()
```

```{r}
# Plot mean of male and mean of female
library(dplyr)
MEAN <- data.selected %>%
  group_by(male,age,cardio,reject) %>%
  summarise(respons = mean(respons))
MEAN
```
```{r}
ggplot(data.selected,aes(x=age,y=respons,color=male)) + geom_line(data=MEAN)
```

```{r}
# Spaghetti Ggplot separated by male =1
p <- ggplot(data=data.selected,aes(x=time,y=respons,group=id))
p <- p + geom_line(col="grey")+stat_summary(aes(group=1),geom="line",fun=mean,linewidth=2)
p + facet_grid(~male,labeller=label_both)

```
```{r}
# Spaghetti Ggplot separated by cardio
p <- ggplot(data=data.selected,aes(x=time,y=respons,group=id))
p <- p + geom_line(col="grey")+stat_summary(aes(group=1),geom="line",fun=mean,linewidth=2)
cardio.labs <- c("Cardio = 0","Cardio = 1")
p + facet_grid(~cardio,labeller = label_both) 
```
```{r}
# Spaghetti Ggplot separated by reject =1
p <- ggplot(data=data.selected,aes(x=time,y=respons,group=id))
p <- p + geom_line(col="grey")+stat_summary(aes(group=1),geom="line",fun=mean,linewidth=2)
p + facet_grid(~reject,labeller=label_both)

```

```{r}
# BoxPlot
ggplot(data.selected,aes(x=as.factor(age),y=respons))+ geom_boxplot(position=position_dodge(1))
```
```{r}
# BoxPlot
ggplot(data.selected,aes(x=as.factor(time),y=respons))+ geom_boxplot(position=position_dodge(1))
```

```{r}
# Box plot by sex
ggplot(data.selected,aes(x=as.factor(time),y=respons,fill=as.factor(male)))+ 
  geom_boxplot(position=position_dodge(1))
```
```{r}
# Box plot by cardio
ggplot(data.selected,aes(x=as.factor(time),y=respons,fill=as.factor(cardio)))+ 
  geom_boxplot(position=position_dodge(1))
```

```{r}
# Box plot by reject
ggplot(data.selected,aes(x=as.factor(time),y=respons,fill=as.factor(reject)))+ 
  geom_boxplot(position=position_dodge(1))
```
#### Spaghetti Plot the response over time with the different persons
```{r}

#data.selected = data[data$id == selected.vector,]# why the dim(data.selected) = 12 x 8
# Plot the respons over time for different id
# If some responses are not available, NA
ggplot(data.selected, aes(x=time, y=respons, group=id,color=id)) + geom_point()  +geom_line()
#ggplot(data.selected, aes(x=time, y=na.pass(respons), group=id,color=id)) + geom_point()  +geom_line()
```

##### Box plot response over time
```{r}
# Box plot
p <- ggplot(data, aes(x=time, y=respons,group =time, color = time)) +  
  geom_boxplot()
p
```
#### Hypothese one
HC level will change with time differently if the REJECT is different
```{r}
#Plot individual data
ggplot(data, aes(x=time, y=respons, group=reject,color=reject)) + geom_point() 
```
##### Box plot 

```{r}
# Box plot
p <- ggplot(data, aes(x=reject, y=respons,group = reject, color = reject)) +  
  geom_boxplot()
p
```


#### Hypothese two
HC level will change with time differently if the gender is different, male (1) has generally higher HC level than female (0)
```{r}
data.male =  data[(data$male == "1" ), ] 
data.female =  data[(data$male == "0" ), ] 
data.male.selected =  data[(data$male == "1" & data$id %in% c(selected)), ] 
data.female.selected =  data[(data$male == "0" & data$id %in% c(selected)), ]
```

#####  Spaghetti plots stratified by variables gender male
```{r}
ggplot(data.male.selected, aes(x=time, y =respons, group=id,color=id)) + geom_point()  +geom_line()+ggtitle("Spaghetti plot for male")+ geom_smooth(method="lm")
ggplot(data.female.selected, aes(x=time, y= respons, group=id,color=id)) + geom_point()  +geom_line()+ggtitle("Spaghetti plot for female")+ geom_smooth(method="lm")
```
##### Box plot
```{r}
p <- ggplot(data, aes(x=time, y=respons,fill= male)) +  
  geom_boxplot()
p a 
```

##### Box plot for male and female
COULD I PLOT THEM IN THE SAME FIGURE?
```{r}

p.box.male <- ggplot(data.male, aes(x=time, y=respons,group = time, color = time)) +  
  geom_boxplot()
p.box.male

p.box.female <- ggplot(data.female, aes(x=time, y=respons,group = time, color = time)) +  
  geom_boxplot()
p.box.female
```



#### Hypothese three
HC level will change with time differently if the age when performing the kidney transplant is younger
```{r}
#Plot individual data
ggplot(data.selected, aes(x=time, y=respons, group=age,color=age)) + geom_point() +geom_line()
```

##### Box plot
```{r}
p <- ggplot(data, aes(x=age, y=respons,group = age, color = age)) +  
  geom_boxplot()
p
```
#### Hypothese four
HC level will change with time differently if the patient has experienced cardio-vascular problem during the years preceding the tranplantation
##### Spathetti plot ?
```{r}
#Plot individual data
ggplot(data.selected, aes(x=time, y=respons, group=cardio,color=cardio)) + geom_point()  +geom_line()
```
##### Bar plot
```{r}
p <- ggplot(data, aes(x=cardio, y=respons,group = cardio, color = cardio)) +  
  geom_boxplot()
p
```

## Correlation analysis for different HC levels along time
For this purpose we need the wide table

```{r}
trenal.wide = trenal[,1:17]
summary(trenal.wide)
```

```{r}
HcCorr = trenal.wide[,c(1:12)]
cor(HcCorr,use="complete.obs" ) # also COV for covariance
library("PerformanceAnalytics")
chart.Correlation(HcCorr,historgram=TRUE)
```


```{r}
cor(trenal.wide$HC0,trenal.wide$HC06)
```
```{r}
# scatter plot matrix

pairs(~HC0+HC06+HC1+HC2+HC3+HC4+HC5+HC6+HC7+HC8+HC9+HC10,data=trenal.wide,
   main="Simple Scatterplot Matrix")
```





```{r}
#Plot individual data
ggplot(data, aes(x=time, y=respons, group=male,color=male)) + geom_point() 
#Spaghetti Plot
ggplot(data, aes(x=time, y=respons, group=male,color=male)) + geom_point() +geom_line()
#Spaghetti with fitted lines
ggplot(data, aes(x=time, y=respons, group=male,color=male)) + geom_point()+ geom_smooth(method="lm",se=F) +geom_line()


```
\newpage
## Linear mixed effect model
```{r}
library(nlme)
library(lme4)
library(lattice)
library(ggplot2)
```

First model only consider fixed effects, no random effects, use linear model and linear mixed model

```{r}
# First model, no random effects
model.noRandomEffects <- lm(respons ~ time,data=data)
summary(model.noRandomEffects)
                      
```
Read from the table, we have a same model for each subject $i$:
$$\textsf{Respons}_i = 37.33685 + 0.26322 \times \textsf{time} + \epsilon_i $$
where $\epsilon$ has a large variance
Residuals:
     Min       1Q   Median       3Q      Max 
-23.3368  -3.8633   0.0393   3.8206  27.1367 
```{r}
# second model random intercept
model.inter <- lmer(respons~time+(1|id),data=data)
summary(model.inter)
```
Read from the table, we have a different model for each subject $i$
$$\textsf{Respons}_i = 37.295 + b_{i0} + (0.2913+b_{i1}) \times \textsf{time}+ \epsilon_i$$
Now the residual $\epsilon_i$ has smaller variance
Scaled residuals: 
    Min      1Q  Median      3Q     Max 
-5.5197 -0.4741  0.0847  0.5683  6.4131 

Is the random intercept giving information?
```{r}
value1 <- as.numeric(2*(logLik(model.inter)-(logLik(model.noRandomEffects))))
value1
p1 = 0.5 * (1-pchisq(value1,1))
p1
```
```{r}
library(nadiv)
#?LRTest
LRTest(logLik(model.inter),logLik(model.noRandomEffects))
```
Thus the model.inter outperforms the linear model

## Third model Random Slope
```{r}
model.slope = lmer(respons ~ time + (time-1|id),data=data)
summary(model.slope)
```
Read from the table, we have a different model for each subject $i$:

$$\textsf{Respons}_i = 37.23239  + (0.35231+b_{i1}) \times \textsf{time}+ \epsilon_i$$

Is this random slope giving information?

```{r}
value2 <- as.numeric(2*(logLik(model.slope)-(logLik(model.noRandomEffects))))
value2
p2 = 0.5*(1-pchisq(value2,1))
p2
LRTest(logLik(model.slope),logLik(model.noRandomEffects))
```
```{r}
value12 <- as.numeric(2*(logLik(model.slope)-(logLik(model.inter))))
value12
p12 = 0.5*(1-pchisq(value12,1))
p12
LRTest(logLik(model.slope),logLik(model.inter))
```

```{r}
value12 <- as.numeric(2*(logLik(model.slope)-(logLik(model.inter))))
value12
p12 = 0.5*(1-pchisq(value12,1))
p12
LRTest(logLik(model.inter),logLik(model.slope))
```
model.inter outperforms than the model.slope

Fourth Model

Fixed effect(changes the same for all subjects) could be time, male, age, reject, cardio
Random effect will start from only intercept, to the slope of gender, age, reject, cardio

## Only fixed effects
```{r}
model.noRandomEffects.full = lm(respons ~ time+age+male+cardio+reject, data=data)
summary(model.noRandomEffects.full)
```

```{r}
data.noNA <- na.omit(data)
summary(data.noNA)

```
## Take the intercept as random effects
```{r}
library(lme4)
model.inter = lmer(respons ~ time+age+male+cardio+reject + (1|id), data=data)
summary(model.inter)
```

```{r}
library(nlme)
model.inter.full = lme(respons ~ time+age+male+cardio+reject, random = ~1|id, data=data.noNA)
summary(model.inter.full)
```
## Take 

Besides we can think which term to consider the fixed effect and random effect
Intercept + slope 1st order term + slope 2nd order term

```{r}
#lme
#data = trenal.long
#lme <- lme(repsons ~ time + age ,data=data)
#lme<-lme(respons~time+age+male+reject+cardio,data=data)
#summary(lme)

#newdata<-data.frame(ID=c(1,2,3,4,5),week=c(3,3,3,3,3))
#newdata$prediction<-predict(lm,newdata=newdata)
#newdata
#predict(lme,newdata=newdata,level=0:1)


```


















