---
title: "Longitudinal Data Analysis "
subtitle: "Case study of Trenal.XLS using Linear Mixed Effect Model"
author: "Group2:Hugo Blain; Oscar Cabanelas;Wanchang Zhang"
date: "`r Sys.Date()`"
output: 
  pdf_document:
  fig_caption: true
  number_sections: yes
  extra_dependencies:
    graphicx: null
    bm: null 
    multirow: null
    multicolumn: null
    hyperref: null
    amsmath : null
bibliography: scholar.bib          
---
\newpage 
\tableofcontents 
\listoffigures
\listoftables
\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE)
```
# Data description

## Backgrounds of the data
The dataset `Trenal.XLS` contains information on patients who received renal graft(kidney transplant). The patients have been followed for at most 10 years.

People with end-stage kidney disease who receive a kidney transplant generally live longer than people with ESRD who are on dialysis. 
However, kidney transplant recipients must remain on immunosuppressants  (medications to suppress the immune system) for the rest of their life to prevent their body from rejecting the new kidney. The long-term immunosuppression puts them at risk for infections and cancer.
Haematocrit level is meassured for each patient who has received renal graft to see if gender, the age to go through the operation, reject or not, cardio history or not will influence the healthy state of a patient after operation.

```{r,echo=FALSE}
library(readxl)
library(knitr)
library(ggplot2)
library(dplyr)
library(lme4)
library(GGally)
library(nlme)
library("PerformanceAnalytics")
```
## Data preprocess

### Import and clean up data `Trenal.XLS`

```{r}
trenal <- read_excel("Trenal.XLS") # summary(trenal)

trenal= trenal[,-18] #remove a noninformative column const

# Continuous or discrete variables
trenal$id = as.factor(trenal$id)
trenal$j = as.factor(trenal$j)
#trenal$time = as.factor(trenal$time)
trenal$male = as.factor(trenal$male)
trenal$cardio = as.factor(trenal$cardio)
trenal$reject = as.factor(trenal$reject)

# Change the name of respons
colnames(trenal)[19] <- "Hc"
trenal.long = trenal[,13:20] # long table form

# Remove j
trenal.long = trenal.long[,-6]
trenal.long.unique <- trenal.long[match( unique(trenal.long$id), trenal.long$id),]# meanHc should replace trenal.long.unique$Hc
trenal.long.noNA <- na.omit(trenal.long)

# Wide table form
trenal.wide = as.data.frame(subset(trenal,trenal$j=="1"))[,1:18] # 1160 x 18
```

### Data Organization
* The input data:
  * id: total `r dim(trenal.long.unique)[1]` persons
  * age to perform the operation: from $15$ to $76$ years old, average is $46.43$ years old
  * male: we observe 494 females and 666 males
  * cardio: 953 persons has experienced a cardio-vascular problem during
the years preceding the transplant, 207 did not.
  * reject: 793 patients shown symptoms of graft rejection during the first
three months after the transportation, 367 has not.

* The response variable Hc level: continous from min $14\%$ to max $65\%$.The Hc level is dependent on the meassured time, individual's age to perform the              operation,  gender, cardio history and reject history.

### Missing Data
```{r,fig.axis=1,fig.align='center'}
# Analyse the NA values descriptively
## First to collect how many NAs are in Hc0, Hc0.5, Hc1, ..., Hc 10
Hc.NA = numeric(12)
for (i in c(1:12)) {
  Hc.NA[i] = sum(is.na(trenal.wide[,i]))
}
# 1   0   1  87 205 314 418 508 595 672 749 812
# The number of missing data / the ideal case have all meassurements for everyone
Hc.NA.percentage = Hc.NA/dim(trenal.long.unique)[1]
missing <- data.frame(rbind(Hc.NA,Hc.NA.percentage))
kable(missing, caption= "Missing data for each measurement",
      col.names =colnames(trenal.wide[c(1:12)]),digits = 3)
plot(Hc.NA.percentage,xaxt="n",xlab ="j",ylab="Missing data%")
axis(side=1,at=c(1,2,3,4,5,6,7,8,9,10,11,12),labels=colnames(trenal.wide)[1:12])

```

Conclusion could be at the first three measurements, there are almost full data
More people tends to miss the measurements when time increases. 
And then we can extract all NA data from the long table to analyse their construction
```{r,fig.align='center'}
trenal.long.NA = trenal.long[is.na(trenal.long$Hc),]
#t = unique(trenal.long.NA$id) # 821 individuals
trenal.long.NA.unique <- trenal.long.NA[match( unique(trenal.long.NA$id), trenal.long.NA$id),]
summary(trenal.long.NA.unique)
```
```{r}
## Conclusion, For the missing data, we can see that 
png(file="MissingValueAnalysis.png",
    width=600, height=1200)
plot.new()
par(mfrow=c(4,2))
## age 
hist(trenal.long.unique$age,title="Age distribution in original data")
hist(trenal.long.NA.unique$age,col="red",title="Age distribution in missing data")

## male
plot(trenal.long.unique$male)
title(main="Gender distribution in original data")
plot(trenal.long.NA.unique$male,col="red")
title(main="Gender distribution in missing data")

## cardio
plot(trenal.long.unique$cardio)
title(main="Cardio distribution in original data")
plot(trenal.long.NA.unique$cardio,col="red")
title(main="Cardio distribution in missing data")

## reject
plot(trenal.long.unique$reject)
title(main="Reject distribution in original data")
plot(trenal.long.NA.unique$reject,col="red") 
title(main="Reject distribution in missing data")
dev.off()
```

The missing data has a similar distribution as the ideal full data set, in age, male, cardio and reject plot. So we may conclude that the missing data are random and not depend on any observed predictors or the response.
# Exploratory Data Analysis

## Univariate summaries

### Plot histogram of continuous variables

```{r,fig.axis=1,fig.align='center'}
hist(trenal.long$Hc,title="Hc distribution in original data")
```

```{r,fig.align='center'}
hist(trenal.long.unique$age,title="age distribution in original data")
```

## Bivariate summaries

### Plot relationship between pairs of variables

```{r,fig.align='center'}
# Bivariate summaries
gg <- ggpairs(data=trenal.long.unique[,2:6])# Here Hc is only one value per individual
gg
```

### Plot the time trend of response 

#### Mean Structure

```{r,fig.axis=1,fig.align='center'}
# To view the mean structure of the Hc for all individuals
ggplot(trenal.long.noNA,aes(x=as.factor(time),y=Hc,group=id))  + geom_line(col="grey")+stat_summary(aes(group=1),geom="line",fun=mean,linewidth=2)+
  labs(title="Line plot of Hc level for all individuals overtime and the mean structure")
```

#### Variance Structure

```{r,fig.axis=1,fig.align='center'}
# To view to variance structure
ggplot(trenal.long.noNA,aes(x=as.factor(time),y=Hc))+ 
  geom_boxplot(position=position_dodge(1))+
  labs(title="Box Plot of Hc level for all indivuduals over time and the variance structure")
```

#### Covariance Structure

```{r,fig.axis=1,fig.align='center'}
HcCorr = trenal.wide[,c(1:12)]
#cor(HcCorr,use="complete.obs" ) # also COV for covariance
chart.Correlation(HcCorr,historgram=TRUE)
```

### Find out a covariate increasing or decreasing the responses time trend#

#### Spaghetti Plot

```{r}
# since the data dimension is large 9551 x 8, we can select random 30 data to have a look 
set.seed(1)
selected <- sample(1:length(unique(trenal.long.noNA$id)),30,replace=T) # random samples and permutations
#selected.vector = as.vector(selected)
data.selected = trenal.long.noNA[(trenal.long.noNA$id %in% c(selected)), ] 
```

* Spaghetti plot group by id

```{r,fig.axis = 1,echo=FALSE}
ggplot(data.selected,aes(x=time,y=Hc,group=id,color=id))+geom_point()+ geom_line()+theme_light()
```

* Spaghetti plot group by male

```{r,fig.axis=1,echo=FALSE}
p <- ggplot(data=trenal.long.noNA,aes(x=time,y=Hc,group=id))
p <- p + geom_line(col="grey")+stat_summary(aes(group=1),geom="line",fun=mean,linewidth=2)
p + facet_grid(~male,labeller=label_both)

```

* Spaghetti plot group by cardio

```{r,echo=FALSE}
# Spaghetti Ggplot separated by cardio
p <- ggplot(data=trenal.long.noNA,aes(x=time,y=Hc,group=id))
p <- p + geom_line(col="grey")+stat_summary(aes(group=1),geom="line",fun=mean,linewidth=2)
cardio.labs <- c("Cardio = 0","Cardio = 1")
p + facet_grid(~cardio,labeller = label_both) 
```

*Spaghetti plot group by reject

```{r, echo=FALSE}
# Spaghetti Ggplot separated by reject =1
p <- ggplot(data=trenal.long.noNA,aes(x=time,y=Hc,group=id))
p <- p + geom_line(col="grey")+stat_summary(aes(group=1),geom="line",fun=mean,linewidth=2)
p + facet_grid(~reject,labeller=label_both)

```


#### Boxplot

* Box plot by male

```{r, echo=FALSE}
ggplot(trenal.long.noNA,aes(x=time,y=Hc,fill=as.factor(male)))+ 
  geom_boxplot(position=position_dodge(1))
```

* Box plot by cardio

```{r,fig.align = "center",echo=FALSE}
# Box plot by cardio
ggplot(trenal.long.noNA,aes(x=as.factor(time),y=Hc,fill=as.factor(cardio)))+ 
  geom_boxplot(position=position_dodge(1))
```

* Box plot by reject

```{r,echo=FALSE}
ggplot(trenal.long.noNA,aes(x=as.factor(time),y=Hc,fill=as.factor(reject)))+ 
  geom_boxplot(position=position_dodge(1))
```

### Data set analysis to see the age effect

```{r}
ggplot(data=trenal.long.noNA,aes(y=Hc,x=age))+geom_point()
```


```{r}
data.groupbyage <- trenal.long.noNA %>%   group_by(age) %>% summarise(avgHc=mean(Hc))
ggplot(data=data.groupbyage,aes(y=avgHc,x=age))+geom_point()
```

## Conclusions after exploring data analysis

* The Hc time trend tends to increase first from 0 to 0.5 year then keep variated during the rest of the meassurements
* The subject related variables age may increase the mean Hc level of a subject
* Male has relative higher Hc level than female
* Cardio or reject play no big difference in the Hc level measurements.


# Multilevel Data Analysis

## Multivariate Linear Model Analysis

```{r}
lm1 <- lm(Hc ~ time, trenal.long)
summary(lm1)

lm2 <- lm(Hc ~ time + age, trenal.long)
summary(lm2)

lm3 <- lm(Hc ~ time + age + male,trenal.long)
summary(lm3)

lm4 <- lm(Hc ~ time + age + male + reject,trenal.long)
summary(lm4)

lm5 <- lm(Hc ~ time + age + male + reject + cardio,trenal.long)
summary(lm5)

anova(lm2,lm3,lm4,lm5)
```

Conclustions:
Variables must keep are: intercept, time, age, male
if p value is 0.01, it is better to add reject
if p value is 0.05, it is better to add cardio 


## Linear Mixed effects Model Analysis

This is inspired from the chapter of <https://bookdown.org/roback/bookdown-BeyondMLR/ch-lon.html>
Longitudinal data is a special example of multilevel data, where

  * Level One is : time and the response variable e.g. Hc level 
  * Level Two is : covariates related to each subject, e.g. age, male, reject, cardio

### Unconditional Means Model to discover variance distribution
We can first try the unconditional Means Model to explore the variance( within subject and between-subject), Define $Y_{ij}$ as the Hc level from subject $i$ and measured time $j$

  * Level One: 
  $$Y_{ij} = a_i + \epsilon_{ij},$$ where $\epsilon_{ij} \sim N(0,\sigma^2)$
  * Level Two: 
  $$a_i = \alpha_0 + u_i,$$ where $u_i \sim N(0,\sigma_{u}^2)$
  
Written in linear mixed effect model is:
 $$Y_{ij} = \alpha_0 + u_i + \epsilon_{ij},$$ where $u_i \sim N(0,\sigma_u^2)$ and $\epsilon_{ij} \sim N(0,\sigma^2)$
 
```{r}
# Model A
library(lme4)
model.a <- lmer(Hc ~ 1 + (1|id),REML=T,data=trenal.long)
summary(model.a)
```


```{r,echo=FALSE}
cat("AIC = ",AIC(model.a),";BIC = ", BIC(model.a))
```

From the output of `model.a`, we obtain estimates of three model parameters:

  * $\hat{\alpha}_0 = 38.16$: the mean of Hc level $\mu_{Hc}$ across all subjects and all years
  * $\hat{\sigma}^2 = 23.07$: the variance in within-subjects deviation, between years of measurements $Hc_j$ and the mean $\mu_{Hc}$ across all subjects and all years
  * $\hat{\sigma_u}^2 = 13.60$: the variance in between-subjects deviation, between subject mean $\mu_{Hc_i}$ and the overall mean $\mu_{Hc}$ across all subjects and all years.
  
The intraclass correlation coefficient:
$$\hat{\rho}= \frac{\hat{\sigma_u}^2}{\hat{\sigma_u}^2+\hat{\sigma}^2} = \frac{13.60}{13.60+ 23.07}=0.371$$

$37.1\%$ of the total variation in Hc levels is attributable to differences among subjects rather than changes over time within each subject.

### Unconditional Growth Model, introducing time in Level One

  * Level One: 
  $$Y_{ij} = a_i + b_i \times time_{ij} + \epsilon_{ij},$$ where $\epsilon_{ij} \sim N(0,\sigma^2)$
  
  * Level Two: 
  
  \begin{align}
  a_i &= \alpha_0 + u_i, \\
  b_i &= \beta_0 + v_i
  \end{align}
  where $\begin{bmatrix} u_i \\ v_i \end{bmatrix} \sim  N\left(\begin{bmatrix} 0 \\ 0 \end{bmatrix}, \begin{bmatrix}\sigma_{u}^2 & \rho_{uv}\sigma_u\sigma_v \\ \rho_{uv} \sigma_u\sigma_v & \sigma_v^2 \end{bmatrix}\right)$


Written in linear mixed effect model is:
 $$Y_{ij} = [\alpha_0 + \beta_0 \times time_{ij}] + [ u_i + v_i \times time_{ij} +  \epsilon_{ij} ],$$ where $\begin{bmatrix} u_i \\ v_i \end{bmatrix} \sim  N\left(\begin{bmatrix} 0 \\ 0 \end{bmatrix}, \begin{bmatrix}\sigma_{u}^2 & \rho_{uv}\sigma_u\sigma_v \\ \rho_{uv} \sigma_u\sigma_v & \sigma_v^2 \end{bmatrix}\right)$ and $\epsilon_{ij} \sim N(0,\sigma^2)$

```{r}
# model b
model.b <- lmer(Hc ~ time + (time|id),REML=T,data= trenal.long)
summary(model.b)
```


```{r,echo=FALSE}
cat("model.a: AIC = ",AIC(model.a),";BIC = ", BIC(model.a),"\n")
cat("model.b: AIC = ",AIC(model.b),";BIC = ", BIC(model.b),"\n")
```
From the `model.b`, we obtain estimates of our six model parameters:
 
* $\hat{\alpha}_0 = 37.2678$: the mean Hc level for the subjects at time 0, $Hc_0$
* $\hat{\beta}_0 = 0.31583$: the mean change in successively measurements during totally $12$ measurements
* $\hat{\sigma}^2 = 20.8883$: the variance in within-subject deviations
* $\hat{\sigma}_u^2 = 13.5731$: the variance between subjects at time 0, $Hc_0$
* $\hat{\sigma}_v^2 = 0.1856$: the variance between subjects in rate of changes in Hc level
* $\rho_{uv} = -0.15$: the correlation in subject's $Hc_0$ and the rate of change in Hc level

The estimated within-subject variance $\hat{\sigma}^2$ decreased by about $9\%$ from the unconditional means model implying that $9\%$ of within-subject variability in Hc level can be explained by a linear increase over time:
$$Pseudo R^2_{L1} = \frac{\hat{\sigma}^2(uncond. means)-\hat{\sigma}^2(uncond. growth)}{\hat{\sigma}^2(uncond. growth)} = \frac{23.07-20.8883}{23.07} = 0.0948$$ 

#### Unconditional growth but only random intercepts

```{r}
model.b1 <- lmer(Hc ~ time + (1|id),REML=T,data= trenal.long)
summary(model.b1)
```


```{r,echo=FALSE}
cat("model.a: AIC = ",AIC(model.a),";BIC = ", BIC(model.a),"\n")
cat("model.b: AIC = ",AIC(model.b),";BIC = ", BIC(model.b),"\n")
cat("model.b1: AIC = ",AIC(model.b1),";BIC = ", BIC(model.b1),"\n")
```

### Modeling other trends over time quadratic

From the spaghetti plots we notice that our Hc level usually increases first very quickly then stays stable. (piecewise linear? )
To reduce the correlation between the linear and quadratic components of time effect, we need to center the time variable first:

```{r}
trenal.long.center <- trenal.long%>%
  mutate(timec = time - 5,timec2 = timec^2)
```


  * Level One: 
  $$Y_{ij} = a_i + b_i \times time_{ij} + c_i \times time_{ij}^2 + \epsilon_{ij},$$ where $\epsilon_{ij} \sim N(0,\sigma^2)$
  
  * Level Two: 
  
  \begin{align}
  a_i &= \alpha_0 + u_i, \\
  b_i &= \beta_0 + v_i, \\
  c_i &= \gamma_0 + w_i,
  \end{align}
  where $\begin{bmatrix} u_i \\ v_i \\w_i \end{bmatrix} \sim  N\left(\begin{bmatrix} 0 \\ 0 \\0 \end{bmatrix}, \begin{bmatrix}\sigma_{u}^2 & \rho_{uv}\sigma_u\sigma_v & \rho_{uw} \sigma_u \sigma_w \\ &  \sigma_v^2 & \rho_{vw} \sigma_v\sigma_w \\ & & \sigma_w^2 \end{bmatrix}\right)$

Written in linear mixed effect model is:
 $$Y_{ij} = [\alpha_0 + \beta_0 \times time_{ij} + \gamma_0 \times time_{ij}^2] + [ u_i + v_i \times time_{ij} + w_i \times time_{ij}^2 +  \epsilon_{ij} ],$$ where  $\begin{bmatrix} u_i \\ v_i \\w_i \end{bmatrix} \sim  N\left(\begin{bmatrix} 0 \\ 0 \\0 \end{bmatrix}, \begin{bmatrix}\sigma_{u}^2 & \rho_{uv}\sigma_u\sigma_v & \rho_{uw} \sigma_u \sigma_w \\ &  \sigma_v^2 & \rho_{vw} \sigma_v\sigma_w \\ & & \sigma_w^2 \end{bmatrix}\right)$ and $\epsilon_{ij} \sim N(0,\sigma^2)$



```{r}
model.c <- lmer(Hc ~ timec + timec2 + (timec + timec2|id),REML=T, data=trenal.long.center)
summary(model.c)
```


```{r,echo=FALSE}
cat("model.a: AIC = ",AIC(model.a),";BIC = ", BIC(model.a),"\n")
cat("model.b: AIC = ",AIC(model.b),";BIC = ", BIC(model.b),"\n")
cat("model.b1: AIC = ",AIC(model.b1),";BIC = ", BIC(model.b1),"\n")
cat("model.c: AIC = ",AIC(model.c),";BIC = ", BIC(model.c),"\n")
```

```{r}
model.c1 <- lmer(Hc ~ timec + timec2 + (1|id),REML=T, data=trenal.long.center)
summary(model.c1)
```


```{r,echo=FALSE}
cat("model.a: AIC = ",AIC(model.a),";BIC = ", BIC(model.a),"\n")
cat("model.b: AIC = ",AIC(model.b),";BIC = ", BIC(model.b),"\n")
cat("model.b1: AIC = ",AIC(model.b1),";BIC = ", BIC(model.b1),"\n")
cat("model.c: AIC = ",AIC(model.c),";BIC = ", BIC(model.c),"\n")
cat("model.c1: AIC = ",AIC(model.c1),";BIC = ", BIC(model.c1),"\n")
```



### Piecewise linear time trend
In the **piecewise linear model**, the complete time span of the study is divided into two segments, with a separate slope relating time to the response in each segment.

* Level One: 
  $$Y_{ij} = a_i + b_i \times time_{1_{ij}} + c_i \times time_{2_{ij}} + \epsilon_{ij},$$ where $\epsilon_{ij} \sim N(0,\sigma^2)$
  
  * Level Two: 
  
  \begin{align}
  a_i &= \alpha_0 + u_i, \\
  b_i &= \beta_0 + v_i, \\
  c_i &= \gamma_0 + w_i,
  \end{align}
  where $\begin{bmatrix} u_i \\ v_i \\w_i \end{bmatrix} \sim  N\left(\begin{bmatrix} 0 \\ 0 \\0 \end{bmatrix}, \begin{bmatrix}\sigma_{u}^2 & \rho_{uv}\sigma_u\sigma_v & \rho_{uw} \sigma_u \sigma_w \\ &  \sigma_v^2 & \rho_{vw} \sigma_v\sigma_w \\ & & \sigma_w^2 \end{bmatrix}\right)$

Written in linear mixed effect model is:
 $$Y_{ij} = [\alpha_0 + \beta_0 \times time_{1_{ij}} + \gamma_0 \times time_{2_{ij}}] + [ u_i + v_i \times time_{1_{ij}} + w_i \times time_{2_{ij}} +  \epsilon_{ij} ],$$ where  $\begin{bmatrix} u_i \\ v_i \\w_i \end{bmatrix} \sim  N\left(\begin{bmatrix} 0 \\ 0 \\0 \end{bmatrix}, \begin{bmatrix}\sigma_{u}^2 & \rho_{uv}\sigma_u\sigma_v & \rho_{uw} \sigma_u \sigma_w \\ &  \sigma_v^2 & \rho_{vw} \sigma_v\sigma_w \\ & & \sigma_w^2 \end{bmatrix}\right)$ and $\epsilon_{ij} \sim N(0,\sigma^2)$
 
In our case study, we can fit separate slope in time $0-0.5$ and $0.5-10$
```{r}
# Modeling piecewise linear time trend with two intervals
time1 = trenal.long$time
time1[time1>0.5] = 0
time2 = trenal.long$time
time2[time2<1] = 0
  
trenal.long.piecewise = trenal.long
trenal.long.piecewise['time1'] <- time1
trenal.long.piecewise['time2'] <- time2

model.b.piecewise <- lmer(Hc ~ time1 + time2 + (1|id),REML=T,data=trenal.long.piecewise)

summary(model.b.piecewise)
```


```{r,echo=FALSE}
cat("model.a: AIC = ",AIC(model.a),";BIC = ", BIC(model.a),"\n")
cat("model.b: AIC = ",AIC(model.b),";BIC = ", BIC(model.b),"\n")
cat("model.b1: AIC = ",AIC(model.b1),";BIC = ", BIC(model.b1),"\n")
cat("model.c: AIC = ",AIC(model.c),";BIC = ", BIC(model.c),"\n")
cat("model.c1: AIC = ",AIC(model.c1),";BIC = ", BIC(model.c1),"\n")
cat("model.b.piecewise: AIC = ",AIC(model.b.piecewise),";BIC = ", BIC(model.b.piecewise), "\n")
```


```{r}
model.b.piecewise1 <- lmer(Hc ~ time1 + time2 + (time2|id),REML=T,data=trenal.long.piecewise)
summary(model.b.piecewise1)
```

```{r,echo=FALSE}
cat("model.a: AIC = ",AIC(model.a),";BIC = ", BIC(model.a),"\n")
cat("model.b: AIC = ",AIC(model.b),";BIC = ", BIC(model.b),"\n")
cat("model.b1: AIC = ",AIC(model.b1),";BIC = ", BIC(model.b1),"\n")
cat("model.c: AIC = ",AIC(model.c),";BIC = ", BIC(model.c),"\n")
cat("model.c1: AIC = ",AIC(model.c1),";BIC = ", BIC(model.c1),"\n")
cat("model.b.piecewise: AIC = ",AIC(model.b.piecewise),";BIC = ", BIC(model.b.piecewise), "\n")
cat("model.b.piecewise1: AIC = ",AIC(model.b.piecewise1),";BIC = ", BIC(model.b.piecewise1), "\n")
```

From the AIC and BIC value, we can see the quadratic model `model.c` outperforms the piecewise linear model `model.b.piecewise1`. However, we believe that in reality the $Hc$ level of a person will not change quadratically. It makes more sense that the Hc level of a patient will increases faster the first half year of his or her operation, and then keep stable with probably some random effects to change over the following years.

With the level one model fixed, we can consider adding level two variables sequentially.

### Adding subject related variable in level two
```{r}
model.b.piecewise.age <- lmer(Hc ~ time1 + time2 + age + (time2|id),REML=T,data=trenal.long.piecewise)
```


```{r,echo=FALSE}
summary(model.b.piecewise.age)
cat("model.a: AIC = ",AIC(model.a),";BIC = ", BIC(model.a),"\n")
cat("model.b: AIC = ",AIC(model.b),";BIC = ", BIC(model.b),"\n")
cat("model.b1: AIC = ",AIC(model.b1),";BIC = ", BIC(model.b1),"\n")
cat("model.c: AIC = ",AIC(model.c),";BIC = ", BIC(model.c),"\n")
cat("model.c1: AIC = ",AIC(model.c1),";BIC = ", BIC(model.c1),"\n")
cat("model.b.piecewise: AIC = ",AIC(model.b.piecewise),";BIC = ", BIC(model.b.piecewise), "\n")
cat("model.b.piecewise1: AIC = ",AIC(model.b.piecewise1),";BIC = ", BIC(model.b.piecewise1), "\n")
cat("model.b.piecewise,age: AIC = ",AIC(model.b.piecewise.age),";BIC = ", BIC(model.b.piecewise.age), "\n")
```


```{r}
model.b.piecewise.male <- lmer(Hc ~ time1 + time2 + male + (time2|id),REML=T,data=trenal.long.piecewise)
```

```{r,echo=FALSE}
summary(model.b.piecewise.male)
cat("model.a: AIC = ",AIC(model.a),";BIC = ", BIC(model.a),"\n")
cat("model.b: AIC = ",AIC(model.b),";BIC = ", BIC(model.b),"\n")
cat("model.b1: AIC = ",AIC(model.b1),";BIC = ", BIC(model.b1),"\n")
cat("model.c: AIC = ",AIC(model.c),";BIC = ", BIC(model.c),"\n")
cat("model.c1: AIC = ",AIC(model.c1),";BIC = ", BIC(model.c1),"\n")
cat("model.b.piecewise: AIC = ",AIC(model.b.piecewise),";BIC = ", BIC(model.b.piecewise), "\n")
cat("model.b.piecewise1: AIC = ",AIC(model.b.piecewise1),";BIC = ", BIC(model.b.piecewise1), "\n")
cat("model.b.piecewise,age: AIC = ",AIC(model.b.piecewise.age),";BIC = ", BIC(model.b.piecewise.age), "\n")
cat("model.b.piecewise,male: AIC = ",AIC(model.b.piecewise.male),";BIC = ", BIC(model.b.piecewise.male), "\n")
```


```{r}
model.b.piecewise.maleage <- lmer(Hc ~ time1 + time2 + male +age +(time2|id),REML=T,data=trenal.long.piecewise)
```


```{r,echo=FALSE}
summary(model.b.piecewise.maleage)
cat("model.a: AIC = ",AIC(model.a),";BIC = ", BIC(model.a),"\n")
cat("model.b: AIC = ",AIC(model.b),";BIC = ", BIC(model.b),"\n")
cat("model.b1: AIC = ",AIC(model.b1),";BIC = ", BIC(model.b1),"\n")
cat("model.c: AIC = ",AIC(model.c),";BIC = ", BIC(model.c),"\n")
cat("model.c1: AIC = ",AIC(model.c1),";BIC = ", BIC(model.c1),"\n")
cat("model.b.piecewise: AIC = ",AIC(model.b.piecewise),";BIC = ", BIC(model.b.piecewise), "\n")
cat("model.b.piecewise1: AIC = ",AIC(model.b.piecewise1),";BIC = ", BIC(model.b.piecewise1), "\n")
cat("model.b.piecewise,age: AIC = ",AIC(model.b.piecewise.age),";BIC = ", BIC(model.b.piecewise.age), "\n")
cat("model.b.piecewise,male: AIC = ",AIC(model.b.piecewise.male),";BIC = ", BIC(model.b.piecewise.male), "\n")
cat("model.b.piecewise,maleage: AIC = ",AIC(model.b.piecewise.maleage),";BIC = ", BIC(model.b.piecewise.maleage), "\n")
```

```{r}
model.b.piecewise.maleagereject <- lmer(Hc ~ time1 + time2 + male +age + reject +(time2|id),REML=T,data=trenal.long.piecewise)
```

```{r,echo=FALSE}
summary(model.b.piecewise.maleage)
cat("model.a: AIC = ",AIC(model.a),";BIC = ", BIC(model.a),"\n")
cat("model.b: AIC = ",AIC(model.b),";BIC = ", BIC(model.b),"\n")
cat("model.b1: AIC = ",AIC(model.b1),";BIC = ", BIC(model.b1),"\n")
cat("model.c: AIC = ",AIC(model.c),";BIC = ", BIC(model.c),"\n")
cat("model.c1: AIC = ",AIC(model.c1),";BIC = ", BIC(model.c1),"\n")
cat("model.b.piecewise: AIC = ",AIC(model.b.piecewise),";BIC = ", BIC(model.b.piecewise), "\n")
cat("model.b.piecewise1: AIC = ",AIC(model.b.piecewise1),";BIC = ", BIC(model.b.piecewise1), "\n")
cat("model.b.piecewise,age: AIC = ",AIC(model.b.piecewise.age),";BIC = ", BIC(model.b.piecewise.age), "\n")
cat("model.b.piecewise,male: AIC = ",AIC(model.b.piecewise.male),";BIC = ", BIC(model.b.piecewise.male), "\n")
cat("model.b.piecewise,maleage: AIC = ",AIC(model.b.piecewise.maleage),";BIC = ", BIC(model.b.piecewise.maleage), "\n")
cat("model.b.piecewise,maleagereject: AIC = ",AIC(model.b.piecewise.maleagereject),";BIC = ", BIC(model.b.piecewise.maleagereject), "\n")
```


#### Comparing nested model using anova

```{r}
drop_in_dev <- anova(model.b.piecewise1,model.b.piecewise.male,test="Chisq")
drop_in_dev
```

```{r}
drop_in_dev <- anova(model.b.piecewise.age,model.b.piecewise.maleage,model.b.piecewise.maleagereject,test="Chisq")
drop_in_dev
```

Finally, our optimal model would be the piecewise linear model with age gender as the levle two variables `model.b.piecewise.maleage`.


$$Y_{ij} = [32.69 + 4.112\times time_{1_{ij}} +0.4066 \times time_{2_{ij}} + 2.347 \times male_i + 0.058747 \times age_i] + [u_i + v_i \times time_{2_{ij}} + \epsilon_{ij} $$
 $$Y_{ij} = [\alpha_0 + \beta_0 \times time_{1_{ij}} + \gamma_0 \times time_{2_{ij}}] + [ u_i + v_i \times time_{1_{ij}} + w_i \times time_{2_{ij}} +  \epsilon_{ij} ],$$ where  $\begin{bmatrix} u_i \\ v_i \end{bmatrix} \sim  N\left(\begin{bmatrix} 0 \\ 0 \end{bmatrix}, \begin{bmatrix}12.038 & -0.3155271 \\ -0.3155271 & 0.1874\end{bmatrix}\right)$,and $\epsilon_{ij} \sim N(0,20.5)$
 
 
```{r}
model.b.piecewise2.maleage <- lmer(Hc ~ time1 + time2 + male +age +(1|id),REML=T,data=trenal.long.piecewise)
```

Considering just keep intercept as random effect
```{r,echo=FALSE}
summary(model.b.piecewise.maleage)
cat("model.a: AIC = ",AIC(model.a),";BIC = ", BIC(model.a),"\n")
cat("model.b: AIC = ",AIC(model.b),";BIC = ", BIC(model.b),"\n")
cat("model.b1: AIC = ",AIC(model.b1),";BIC = ", BIC(model.b1),"\n")
cat("model.c: AIC = ",AIC(model.c),";BIC = ", BIC(model.c),"\n")
cat("model.c1: AIC = ",AIC(model.c1),";BIC = ", BIC(model.c1),"\n")
cat("model.b.piecewise: AIC = ",AIC(model.b.piecewise),";BIC = ", BIC(model.b.piecewise), "\n")
cat("model.b.piecewise1: AIC = ",AIC(model.b.piecewise1),";BIC = ", BIC(model.b.piecewise1), "\n")
cat("model.b.piecewise,age: AIC = ",AIC(model.b.piecewise.age),";BIC = ", BIC(model.b.piecewise.age), "\n")
cat("model.b.piecewise,male: AIC = ",AIC(model.b.piecewise.male),";BIC = ", BIC(model.b.piecewise.male), "\n")
cat("model.b.piecewise,maleage: AIC = ",AIC(model.b.piecewise.maleage),";BIC = ", BIC(model.b.piecewise.maleage), "\n")
cat("model.b.piecewise2.maleage: AIC = ",AIC(model.b.piecewise2.maleage),";BIC = ", BIC(model.b.piecewise2.maleage), "\n")
```

```{r}
drop_in_dev <- anova(model.b.piecewise.age,model.b.piecewise.maleage,model.b.piecewise2.maleage,test="Chisq")
drop_in_dev
```

### Two Stage Model Analysis comparison

```{r}
trenal.grouped<-groupedData(Hc~time1+time2|id,trenal.long.piecewise, inner = ~ male+cardio+reject, labels=list(y="level of Haematocrit"),units=list(y="%"))
modlist1 <- lmList (Hc ~ time1+time2|id, trenal.grouped , na.action = na.pass)
#confint(lmList)
a = coef(modlist1)[,1]
b = coef(modlist1)[,2]
c = coef(modlist1)[,3]
plot.new()
par(mfrow=c(1,3))
hist(a,xlab="intercepts",ylab="Frequency",title="Histogram for intercepts")
hist(b,xlab="slopes for t1",ylab="Frequency",title="Histogram for time1 slopes")
hist(c,xlab="slopes for t2",ylab="Frequency",title="Histogram for time2 slopes")
```

Then step two fit the `a,b,c` to the lm analysis with variables `age male `

```{r}
# appendix the value of a, b, c to the original data frame
abc <- data.frame(id=as.numeric(attributes(modlist1)$names),
                 a=coef(modlist1)[,1],b=coef(modlist1)[,2],c=coef(modlist1)[,3])
abc <- merge(trenal.long.piecewise, abc[,c("id","a", "b","c")], by.x = "id", all.x = T)
abc <- abc[order(abc[,1]),]
head(abc)
```
```{r}
### Fit a model for the intercept
fita<-lm(a~ age +male  ,abc)
summary(fita)
confint(fita)

### Fit a model for the slope t1

fitb<-lm(b ~ age + male ,abc)
summary(fitb)
confint(fitb)

### Fit a model for the slope t2

fitc<-lm(c~ age + male ,abc)
summary(fitc)
confint(fitc)

```
We can summarise the two stage model result as


* Level One: 
  $$Y_{ij} = a_i + b_i \times time_{1_{ij}} + c_i \times time_{2_{ij}} + \epsilon_{ij},$$ where $\epsilon_{ij} \sim N(0,\sigma^2)$
  
  * Level Two: 
  
  \begin{align}
  a_i &= 32.607514 + 0.068853 \times age_i + \alpha_2 \times male_i + u_i, \\
  b_i &= \beta_0 + \beta_1 \times age_i + \beta_2 \times male_i + v_i, \\
  c_i &= \gamma_0 + \gamma_1 \times age_i + \gamma_2 \times male_i + w_i,
  \end{align}
  where $\begin{bmatrix} u_i \\ v_i \\w_i \end{bmatrix} \sim  N\left(\begin{bmatrix} 0 \\ 0 \\0 \end{bmatrix}, \begin{bmatrix}\sigma_{u}^2 & \rho_{uv}\sigma_u\sigma_v & \rho_{uw} \sigma_u \sigma_w \\ &  \sigma_v^2 & \rho_{vw} \sigma_v\sigma_w \\ & & \sigma_w^2 \end{bmatrix}\right)$

The simplified linear mixed effect model is:
\begin{equation}  Y_{ij} = [\alpha_0 + \beta_0 \times time_{1_{ij}} + \gamma_0 \times time_{2_{ij}}] + [ u_i + v_i \times time_{1_{ij}} + w_i \times time_{2_{ij}} +  \epsilon_{ij} ], \end{equation}
 where  $\begin{bmatrix} u_i \\ v_i \end{bmatrix} \sim  N\left(\begin{bmatrix} 0 \\ 0 \end{bmatrix}, \begin{bmatrix}12.038 & -0.3155271 \\ -0.3155271 & 0.1874\end{bmatrix}\right)$,and $\epsilon_{ij} \sim N(0,20.5)$
 
 
```{r}
summary(fita)
```




* Level One: 
  $$Y_{ij} = a_i + b_i \times time_{1_{ij}} + c_i \times time_{2_{ij}} + \epsilon_{ij},$$ where $\epsilon_{ij} \sim N(0,\sigma^2)$
  
  * Level Two: 
  
  \begin{align}
  a_i &= \alpha_0 + \alpha_1 \times age_i +\alpha_2 \times male_i  + u_i, \\
  b_i &= \beta_0 + \beta_1 \times age_i + \beta_2 \times male_i  + v_i, \\
  c_i &= \gamma_0 + \gamma_1 \times age_i + \gamma_2 \times male_i + w_i,
  \end{align}
  where $\begin{bmatrix} u_i \\ v_i \\w_i \end{bmatrix} \sim  N\left(\begin{bmatrix} 0 \\ 0 \\0 \end{bmatrix}, \begin{bmatrix}\sigma_{u}^2 & \rho_{uv}\sigma_u\sigma_v & \rho_{uw} \sigma_u \sigma_w \\ &  \sigma_v^2 & \rho_{vw} \sigma_v\sigma_w \\ & & \sigma_w^2 \end{bmatrix}\right)$

The simplified linear mixed effect model is:
\begin{align}Y_{ij} &= [\alpha_0 + \alpha_1\times age_i + \alpha_2 \times male_i \\ & +  \beta_0 \times time_{1_{ij}} + \beta_1 \times age_i \times time_{1_{ij}} + \beta_2 \times male_i \times time_{1_{ij}} \\ &+ \gamma_0 \times time_{2_{ij}} + \gamma_1 \times age_i \times time_{2_{ij}} +  \gamma_2 \times male_i \times time_{2_{ij}} ] \\ & + [u_i + v_i \times time_{1_{ij}} + w_i \times time_{2_{ij}}+\epsilon_{ij}]\end{align}
where $\begin{bmatrix} u_i \\ v_i \\w_i \end{bmatrix} \sim  N\left(\begin{bmatrix} 0 \\ 0 \\0 \end{bmatrix}, \begin{bmatrix}\sigma_{u}^2 & \rho_{uv}\sigma_u\sigma_v & \rho_{uw} \sigma_u \sigma_w \\ &  \sigma_v^2 & \rho_{vw} \sigma_v\sigma_w \\ & & \sigma_w^2 \end{bmatrix}\right)$


#### The equivalent Linear Mixed Effects model should be defined as
```{r}
model.b.piecewise.equivalent <- lmer(Hc ~ age + male + time1+ age:time1+male:time1+time2+age:time2+male:time2+(time1+time2|id),REML=T,data=trenal.long.piecewise)
t = fixef(model.b.piecewise.equivalent)
  
```

```{r,echo=FALSE}
summary(model.b.piecewise.equivalent)
cat("model.a: AIC = ",AIC(model.a),";BIC = ", BIC(model.a),"\n")
cat("model.b: AIC = ",AIC(model.b),";BIC = ", BIC(model.b),"\n")
cat("model.b1: AIC = ",AIC(model.b1),";BIC = ", BIC(model.b1),"\n")
cat("model.c: AIC = ",AIC(model.c),";BIC = ", BIC(model.c),"\n")
cat("model.c1: AIC = ",AIC(model.c1),";BIC = ", BIC(model.c1),"\n")
cat("model.b.piecewise: AIC = ",AIC(model.b.piecewise),";BIC = ", BIC(model.b.piecewise), "\n")
cat("model.b.piecewise1: AIC = ",AIC(model.b.piecewise1),";BIC = ", BIC(model.b.piecewise1), "\n")
cat("model.b.piecewise,age: AIC = ",AIC(model.b.piecewise.age),";BIC = ", BIC(model.b.piecewise.age), "\n")
cat("model.b.piecewise,male: AIC = ",AIC(model.b.piecewise.male),";BIC = ", BIC(model.b.piecewise.male), "\n")
cat("model.b.piecewise,maleage: AIC = ",AIC(model.b.piecewise.maleage),";BIC = ", BIC(model.b.piecewise.maleage), "\n")
cat("model.b.piecewise2.maleage: AIC = ",AIC(model.b.piecewise2.maleage),";BIC = ", BIC(model.b.piecewise2.maleage), "\n")
cat("model.b.piecewise.equivalent: AIC = ",AIC(model.b.piecewise.equivalent),";BIC = ", BIC(model.b.piecewise.equivalent), "\n")
```

```{r}

coeff.twostage = c(extract_numeric(fita$coefficients), extract_numeric(fitb$coefficients), extract_numeric(fitc$coefficients))

fixeff = fixed.effects(model.b.piecewise.equivalent)
fixeff.new = fixeff
fixeff.new[5] <-  fixeff[6]
fixeff.new[6] <-  fixeff[7]
fixeff.new[7] <-  fixeff[5]

coeff.LMM = extract_numeric(fixeff.new)

#coefficients.name = #c('$\alpha_0$','$\alpha_1$','$\alpha_2$','$\beta_0$','$\beta_1$','$\beta_2$','$\gamma_0$','$\gamma_1$','$\gamma_2$')

coefficients.name = c('alpha_0','alpha_1','alpha_2','beta_0','beta_1','beta_2','gamma_0','gamma_1','gamma_2')


# create matrix with 4 columns and 4 rows
coefficients.comparison= cbind(coefficients.name,coeff.twostage,coeff.LMM)

# specify the column names and row names of matrix
#colnames(data) = c('$\alpha_0$','$\alpha_1$','$\alpha_2$','$\beta_0$','$\#beta_1$','$\beta_2$','$\gamma_0$','$\gamma_1$','$\gamma_2$')
#colnames(data) <- c('alpha_0','alpha_1','alpha_2','beta_0','beta_1','beta#_2','gamma_0','gamma_1','gamma_2')

#rownames(data) <- c('row1','row2','row3','row4')

# assign to table
#final=as.table(data)

knitr::kable(coefficients.comparison)

```




































`

